### 2024-12-19 Notes/Notes.tex
Correlation factor results

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage}
\usepackage[top=1.5cm, bottom=3.5cm, left=2.2cm, right=2.2cm]{geometry}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd, esint}
\usepackage{lastpage}
\usepackage{enumerate}
\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[english]{babel}
\usepackage{lipsum}
\usepackage[table,xcdraw]{xcolor}
\usepackage{enumitem}
\usepackage{float}
\usepackage{chemfig}
\usepackage{yfonts}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multicol}
\useunder{\uline}{\ul}{}

@@

\end{figure}
\begin{definition}\label{definitionPersonCoefficient}
Pearson correlation coefficient: Given set of pairs of data $\{(x_i,y_i)\}_{i=1}^n$ the pearson correlation factor $r_{xy}$ is defined as,
\begin{align}
r_{xy}=\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum_{i=1}^n(x_i-\bar{x}^2)}\sqrt{\sum_{i=1}^n(y_i-\bar{y}^2)}},
\end{align}
where $\bar{x}$ and $\bar{y}$ stand for the average value of ${x_i}_{i=1}^n$ and ${y_i}_{i=1}^n$ respectively. Note that $r_{xy}\in[-1,1]$. Therefore the sign of $r_{xy}$ is tightly related to the sign of alinear regression, more precisely if $x>0$, "$y$" generally\footnote{We would like to remark that the word "generally" stands for "the majority of the cases", since "generally" is commonly interpreted as a non-scientific/non-objective word} increases when "$x$" increases, as well as if $x<0$, "$y$" decreases when "$x$" increases.

@@

\begin{observation}
We have extracted all the aviable data from the \emph{ChEMBL} database with a known $IC_{50}$ value for the COX-2. After deleting  entries with the same canonical smile we remain with 4827 entries. We filter, keeping just those with a $IC_{50}\in [-1,200]$\footnote{Note that the lower constrain won't be ever surpassed by the $IC_{50}$ defintion \ref{definitionIC50}. This is equivalent to avoid constraining the data.}. Then we obtain a remaining set of 1518 entries represented in the Figure (\ref{figDataRepresentationFiltering}).
\begin{figure}[H]
\centering
\includegraphics[width = 0.5\textwidth]{../Plots/DataGapRepresentationIC50percentageEresed20.0Gap_min_17Gap_max_47.pdf}
\caption{Data representation after the filtering.}
\label{figDataRepresentationFiltering}
\end{figure}
\end{observation}

@@

\begin{observation}
Once the chemical descriptors are computed, we proceed to calculate the Pearson correlation coefficients as defined in Definition (\ref{definitionPersonCoefficient}). The results are subsequently plotted in Figure (\ref{figPersonCorrelationFactors}). From the plot, we observe a “Gaussian bell” shape centered around zero, which aligns with our expectations. This suggests that there is no inherently preferred set of chemical descriptors for the protein, a finding that may initially seem at odds with the project’s goals.

However, our hypothesis states that \emph{there exists a combination (or combinations) of chemical descriptors that are directly correlated with the inhibition of the protein}. This observation does not contradict that statement. Intuitively, we believe that filtering out descriptors with the lowest  $r^2$  values might lead to improved results. This possibility will be explored in subsequent discussions.
\begin{figure}[H]
\centering
\includegraphics[width = 0.5\textwidth]{../Plots/CorrelationFactors.pdf}
\caption{Representation of the Pearson correlation factors.}
\label{figPersonCorrelationFactors}
\end{figure}

\end{observation}
\end{multicols}
\newpage

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage}
\usepackage[top=1.5cm, bottom=3.5cm, left=2.2cm, right=2.2cm]{geometry}
\usepackage{amsmath,amsthm,amsfonts,amssymb,amscd, esint}
\usepackage{lastpage}
\usepackage{enumerate}
\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[english]{babel}
\usepackage{lipsum}
\usepackage[table,xcdraw]{xcolor}
\usepackage{enumitem}
\usepackage{float}
\usepackage{chemfig}
\usepackage{yfonts}
\usepackage{tikz}
\usepackage{forest}
\usepackage{wrapfig}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multicol}
\useunder{\uline}{\ul}{}

@@

\end{figure}
\begin{definition}\label{definitionPersonCoefficient}
Pearson correlation coefficient: Given set of pairs of data $\{(x_i,y_i)\}_{i=1}^n$ the pearson correlation factor $r_{xy}$ is defined as,
\begin{align}
r_{xy}=\frac{\sum_{i=1}^n(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum_{i=1}^n(x_i-\bar{x}^2)}\sqrt{\sum_{i=1}^n(y_i-\bar{y}^2)}},
\end{align}
where $\bar{x}$ and $\bar{y}$ stand for the average value of ${x_i}_{i=1}^n$ and ${y_i}_{i=1}^n$ respectively. Note that $r_{xy}\in[-1,1]$. Therefore the sign of $r_{xy}$ is tightly related to the sign of alinear regression, more precisely if $x>0$, "$y$" generally\footnote{We would like to remark that the word "generally" stands for "the majority of the cases", since "generally" is commonly interpreted as a non-scientific/non-objective word} increases when "$x$" increases, as well as if $x<0$, "$y$" decreases when "$x$" increases.
\end{definition}
\begin{definition}\label{definitionDecisionTree}
A decision tree is a classification algorithm based on a series of ordered \emph{if} statements. The algorithm begins at the top of the tree, where a question is posed to the data. Depending on the answer, the data follows different branches, each corresponding to a subsequent question. This process is repeated at each node until the data reaches the bottom of the tree, where the path it has followed determines the classification of the given data. \par

For example, a simple depiction of a decision tree is shown in Figure (\ref{figExDecisionTree}), where the decision of whether to enjoy a golfing day or not is determined based on a series of questions.
\begin{figure}[H]
\begin{forest}
  for tree={ draw, rounded corners, align=center, edge={->}, parent anchor=south, child anchor=north, s sep+=5mm, l sep+=5mm  }
  [Is it Sunny?
    [$\checkmark$\\ Is Humidity High? 
      [$\checkmark$\\ Don't Play, draw=red] 
      [$\times$\\ Play, draw=green] 
    ]
    [$\times$\\ Is it Rainy? 
      [$\times$\\ Is it Windy? 
        [$\checkmark$\\ Don't Play, draw=red] 
        [$\times$\\ Play, draw=green] 
      ]
      [$\checkmark$\\ Don't play, draw=red] 
    ]
  ]
\end{forest}
\caption{Example of a decision tree}
\label{figExDecisionTree}
\end{figure}
This kind of algorithm is vividly present in the chemical landscape, for instance in the spectroscopy realm the determination of a molecule's symmetry group is provided by a decision tree such as Figure (\ref{figExScpectroDecisionTree})

\begin{figure}[H]
\centering
\includegraphics[width = 0.5\textwidth]{GeneralSources/Point_group_determination_flowchart_v2.png}
\caption{Decision tree for determining the point group of a molecule}
\label{figExScpectroDecisionTree}
\end{figure}

@@

\begin{observation}
We have extracted all the aviable data from the \emph{ChEMBL} database with a known $IC_{50}$ value for the COX-2. After deleting  entries with the same canonical smile we remain with 4827 entries. We filter, keeping just those with a $IC_{50}\in [-1,200]$\footnote{Note that the lower constrain won't be ever surpassed by the $IC_{50}$ defintion \ref{definitionIC50}. This is equivalent to avoid constraining the data.}. Then we obtain a remaining set of 1518 entries represented in the Figure (\ref{figDataRepresentationFiltering}).
\begin{figure}[H]
\centering
\includegraphics[width = 0.5\textwidth]{../Plots/DataGapRepresentationIC50percentageEresed25.0Gap_min_15Gap_max_50.pdf}
\caption{Data representation after the filtering.}
\label{figDataRepresentationFiltering}
\end{figure}
\end{observation}

@@

\begin{observation}\label{observationCorrelationFactor}
Once the chemical descriptors are computed, we proceed to calculate the Pearson correlation coefficients as defined in Definition (\ref{definitionPersonCoefficient}). The results are subsequently plotted in Figure (\ref{figPersonCorrelationFactors}). From the plot, we observe a “Gaussian bell” shape centered around zero, which aligns with our expectations. This suggests that there is no inherently preferred set of chemical descriptors for the protein, a finding that may initially seem at odds with the project’s goals.

However, our hypothesis states that \emph{there exists a combination (or combinations) of chemical descriptors that are directly correlated with the inhibition of the protein}. This observation does not contradict that statement. Intuitively, we believe that filtering out descriptors with the lowest  $r^2$  values might lead to improved results. This possibility will be explored in subsequent discussions.
\begin{figure}[H]
\centering
\includegraphics[width = 0.5\textwidth]{../Plots/ChEMBL_ExtractorData_CHEMBL372_IC50_1000CorrelationFactors.pdf}
\caption{Representation of the Pearson correlation factors.}
\label{figPersonCorrelationFactors}
\end{figure}
\end{observation}

\begin{observation}
The first algorithm implemented in this project is the \emph{Random Forest} machine learning algorithm. Although the project’s primary focus is not the detailed workings of machine learning algorithms, it is natural to ponder the essence of what the machine is doing. The name “Random Forest” is directly tied to the analogy between a tree and a decision tree. \par

As described in Definition (\ref{definitionDecisionTree}), the Random Forest algorithm generates a collection of random decision trees. The number of trees in the forest is a configurable parameter of the algorithm. More specifically, the Random Forest algorithm produces a set of decision trees, each trained on a random subset of the training data. The structure of each tree adjusts during training to make predictions that closely approximate the true target values. The final output of the Random Forest is typically an aggregate of the predictions from all the individual trees, such as a majority vote for classification tasks or an average for regression tasks. \par

One notable characteristic of this method is its inherent randomness. Each Random Forest model is unique, even if configured with the same parameters and initialized under identical conditions. However, experimental evidence suggests that this randomness does not significantly impact the results. If two Random Forest algorithms are trained on the same dataset using identical parameters, their outputs will converge, even if the internal structures of their decision trees differ.
\end{observation}

\begin{observation}
From Observation (\ref{observationCorrelationFactor}), one might hypothesize that training with chemical descriptors that have higher correlation factors, or discarding those with low correlation factors, would lead to better predictions. Numerically, however, this hypothesis is contradicted, as shown in Table (\ref{tableCorrelationFactorVariation}). Despite this, the result remains consistent with our core hypothesis: the correlation factor of a chemical descriptor does not necessarily indicate its relevance in protein inhibition. Our key assertion is that \emph{there exists a combination (or combinations) of chemical descriptors that are directly correlated with the inhibition of the protein}. \par

This suggests that the specific combinations of descriptors correlated with protein inhibition are not dependent on the individual correlation factor values. At first glance, a lack of correlation does not imply a lack of relationship to the inhibition!

\end{multicols}
\begin{table}[H]
\centering
\caption{Results for; Maximum $IC_{50}$ = 200 nM, Gap of the 25.0\% of the set, the 20.0\% of the remaining data is saved for testing and the  80.0\% for training. The training is done with 200 trees.}
\label{tableCorrelationFactorVariation}
\begin{tabular}{c||c|c|c|c}
          &$\sigma\leq0$ & $\sigma\leq0.05$ & $\sigma\leq0.075$  & $\sigma\leq0.10$ \\\hline\hline
$r^2$ & 0.9607020297015885 & 0.9608545198149474 & 0.9605224652065423 & 0.9599158186858139 \\\hline
Mean Squared Error & 162.70220420738363 & 162.0708617391865 & 163.4456405456867 & 165.9572900112382 \\\hline
number of descriptors & 2911 & 1485 & 1001 & 568
\end{tabular}
\end{table}
\newpage
